Efficiency

### 2018-01-08 DraftExec.Rnw
merge?

The development of the full substance of this paper, has come about due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we shared had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification, and confidence, to meaningfully inform debate.

@@

Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.  In an idealised circumstance, the flow rate increases directly with vessel speed.  Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.  However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.  So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{figLPumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.  One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.

@@

Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.  They simply directly impact the total amount of energy available for consumption before recharging batteries while snorkeling. There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.  The nominal impact, however increases directly and linearly with the total energy stored.

The results in nominal terms are striking in their magnitude.  In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.  Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.  Even under the eficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type214 operating without AIP.

The in terms of submerged range, the same proportional impact is also extremely significant.  Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.  That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.  The tactical significance of such distances are not likely to be lightly dismissed. If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.  This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.  It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.

@@

Only brief is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass. The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.  Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.  This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame. This would plausibly (under the central line selections) amount to more than month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.  This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.

@@

\subsection{The interaction of hotel load, battery energy density, and efficiency curves}

@@

\subsection{The importance of cavitation inception}
It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).  Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system, if it were to occur, owing in particular to the low freqency of diesel engines.

Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be vastly smaller in magnitude.  This owes to the essentially incompressible nature of water.  Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.  All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.  This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.  It also doesn't hold true within a turbulent flow underwater.  If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.  But at a distance, these fluctuations tend to substantially cancel out.  The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.  Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.

However, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.  This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in adegradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.  As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarines 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.

@@

Whilst a range of very particular effects may be at work, perhaps the largest trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.  The blade velocity relative to the oncoming stream is also a close indicator, which will be closely related for an open propeller.  Consequently, since the thrust required for self-propulsion increases with velocity squared, the propensity for cavitation occur will necessarily tend to follow this relationship also.  As such, modest changes in speed can have a big impact on the degree of probability of cavitation occurring. Halving your speed will roughly reduce local pressure fluctions by a factor of four. Consequently, even for non-specialised propellers, there is generaly some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.

\subsection{The impact of depth on cavitation}
It is crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.  As such, the speed at which cavitation starts to occur, will necessarily be a function of depth.  Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.  As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).  Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring \ref{fig:InceptionDepth.png}.

@@

The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model. As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.  However at much higher speeds, sine the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.

@@

It is noteworthy also that the vapour pressure of water is an important component in the equations that describe the conditions for cavitation to occur.  It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs.  Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on likelihood of cavitation occurring.  However, at plausible temperatures the impact is small.  The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilopascals.  Diving just 10m deeper adds over 100 kilopascals to static pressure.  In many circumstances, altering your speed by a couple of knots will have an even larger impact on local pressure reductions.  Consequently, the impact of temperature on the inception of cavitation is essentially negligible.  The difference between inception contours in 20 degree and 30 degree water are given by the difference between the black and red-dashed contours in Figure \ref{fig:InceptionContours}.

\subsection{The significance of flow separation}
The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at very low speeds which correspond to high advance ratios when load remains constant.  Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.  This effect is well known and document.  A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.

@@

Whilst at very low power levels, or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.  However, given the inherently unpredictable and unsteady nation of such vortex formation and shedding, the this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.  In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.  This would necessarily increase the overall levels of noise generated.  It is plausible that the noise is of sufficiently small level that it is not considered a significant compromise of stealth.  In essence, the higher levels of noise lie below some threshold of significance.  However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.

@@

It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.  Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).  Changes in density which are gradual, often representing soft, squishy, or broken surfaces tend, to have the effect of absorbing sound, where the acoustic energy is converted to heat or degradation of the material, rather than reflecting it.  Given the significance of increased drag on any surfaces within the propulsor, it is highly unlikely that any attempt at damping is made within the pumpjet.  However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.  Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelenght, leading to increased thickness of the transition substance or damping material.  Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.

@@

It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.  Constructing especially thin layers, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.  Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.

The degree to which the total noise generated is reflected internally will also be shape-dependent.  A more concave inner surface, consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.  Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.

@@

When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.  However, at very low speeds this effect would be greatly diminished.

It should be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.  This phenomenon generally serves to make the art of underwater acoustics extremely complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.  However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.

@@

This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding. In the absence of flow separation, and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.  This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.  The steel shroud would be very effective in reflecting such high frequency noise.  Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first. These sounds might also be very effectively reflected by the shroud, and consequently diminished.  However, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence generate noise of a much lower frequency.  In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.  Further to this effect, the rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature, of even lower frequency.  Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.

@@

The foregoing consideration suggests that the acoustic signature of a jet-powered submarine might not reduce steadily with speed all the way to stationary movement.  It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.  However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.  It is also possible that the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low as to not be of significant tactical consequence.  However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.

\section{Conclusions}
This is a pretty bad result for the pumpjet.

>>>>

The development of the full substance of this paper has come about due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.

@@

Charts such as these are generally used with dimensionless ratios on the axes (such as $\Delta V / V_0$) to allow the most general solutions to be used to solve for any given particular vessel size, speed, or loading requirement which might be required, these  confusing to many readers unfamiliar with hydrodynamics.  For the avoidance of any confusion, it is important to note that the value for $\Delta V / V_0$ for a given vessl and propulor may tend to increase with increasing waterspeed of the vessel, however one cannot rely upon this particular value diminishing towards zero at very low speeds, as it does for related by different metrics.  In an idealised circumstance, the flow rate increases directly with vessel speed.  Given that the force produced scales with $\Delta V$, which at constant values of $\Delta V / V_0$ increase in proportion to (if, in the idealised circumstance, vessel speed is equal to $V_0$), an increase in thrust can theoretically match the $V_0^2$ requirement imposed by drag without any change in $\Delta V / V_0$.  However, in practice the water encountered by the propulsor is slowed by the wake field, and so the drag encountered by the vessel may increase faster than velocity of the water entering the intake, an effecnt which is likely non-linear with vessel speed.  So whilst the trend with regards to vessel speed can be known with regards to the graphs in \cite{fig:PumpjetEfficiency}, a direct translation of the x-axis used to vessel speed at any given point is difficult prove without additional system information.  One might generally infer that the sorts of declines in efficiency observed as vessel speed discussed in recent literature correspond with movement towards very low values of $\Delta V / V_0$, though it is possible that alternative derivations have been used, or that the phenomenon known to occur in practice (such as flow separation) aren't adequately captured by such ideal equations derived from momentum theory in order to make a direct translation into real-world application.

@@

Changes in the effective battery energy density will have the same effect as having larger masses of battery stored.  They equivalently impact the total amount of energy available for consumption before recharging batteries while snorkeling. There is therefore no shift in the proportional impact on range and endurance due to this factor, as can be seen in the lower graphs in Figure \ref{fig:sens_battery}.  The nominal impact, however increases directly and linearly with the total energy stored.

The results in nominal terms are striking in their magnitude.  In terms of enduracy, doubling the stored electrical energy, for example by adopting lithium batteries, increase the \textit{difference} between the two systems under the central efficiency curve assumptions to nearly four days (93 hours) at 4kt.  Given that most modern, capable submarines are estimated to only be able to achieve dived endurances between 60 and 100 hours on batteries at 4kt \parencite{buckingham2008submarine} this cannot be dismissed in terms of its tactical significance.  Even under the efficiency curve assumptions representing the narrowest spearation considered, the difference in this case would be 59 hours, almost equal to the estimated dived endurance of a Type-214 operating without AIP.

In terms of submerged range, the same proportional impact is also extremely significant.  Doubling the energy stored inceases the \textit{difference} between two otherwise identical submarines to over 400nm under central assumptions at 4.5kt.  That's a larger distance than the width of the Java Sea, the most central body of water in the Indonesian Archipelago, or most of the length of the Malacca Strait.  The tactical significance of such distances are not likely to be lightly dismissed. If far smaller amounts of energy are embarked, the nominal differences reduce to perhaps 100nm, in direct proportion to the reduction of the submarine's dived endurance.  This is still a non-trivial distance in terms of it's tactical signicance, and comprises a substantial fraction of the dived ranges of most non-AIP submarines.  It still represents a significant difference in operational terms, equal approximately to the width of Hainan Island, or neraly twice the length of the Sunda Strait.

@@

Only brief consideration is given here to the impact of Air Independent Propulsion, which has substantially higher energy densities possible even than Lithium, and also amounts to the addtional embarkation of greater mass. The volumetric energy density of the feedstocks to modern AIP systems (liquid oxygen and methanol) is estimated to be approximately ten times greater than that of lead-acid batteries \parencite{davies2016}, around 4.5MJ/L.  Given the greater mass embarked, and additional studies suggesting that at constant speeds they might extend the dived endurance of a modern submarine from around 60 hours to over 1000 hours (a factor of roughly sixteen) \parencite{buckingham2008submarine} an approximate estimation of the the direction and magnitude of this impact may by scaling up energy density by this factor.  This would amount to adding three further frames to the right of Figure \ref{fig:sens_battery}, with the energy density being doubled in each successive frame. This would plausibly (under the central line selections) amount to more than a month of dived endurance \textit{difference} between AIP inclusive submarines with propellers or pumpjets, amounting to thousands of nautical miles of continuously submerged travel.  This difference would also now be material in terms of the overall range of the submarine amounting to a further 15-20\% reduction in the submarine's overall range.

@@

\subsection{The interaction of hotel load, battery energy density, and efficiency curves}
It is interesting to observe how the the combination of changes to the hotel load and battery density interact when altered together, as shown in Figure \ref{fig:sens_corners}.  To the extent that each of these effects contribute to increasing or decreasing the relative difference between the systems, the two clearly combine in a multiplicative fashion.

Where battery density and hotel load are both reduced, the proportional change has a much higher and sharper peak at around 5kt, reaching an 83\% peak at 4.5kt under central assumptions.  For high hotel load and battery density, the porportional impact is lower at a higher speed, at 43\% at 7kt. However, it is intereting to note that the nominal impact of each of these different scenarios is somewhat more similar in magnitude for both endurance and range.  For low hotel load and battery store, the endurance difference peaks at 54 hours at 3 kt, and range difference peaks at 190nm at 4kt.  For raised hotel loads and battery stores, the difference is shifted to higher speeds and broadened somewhat, with a difference of 39hours at 4-5kt, and 200nm at 5-6kt.

The most striking feature of this figure however is the impact on the submarine's performance when hotel load is lowered, and battery density is raised simultaneously.  Here the two factors combined have a proportional impact of around 80\% improvement for the switch to propellers under central assumptions, and over 150\% in the scenario of widest separation.  However, in nominal terms, this impact is even more significant, with central scenarios showing a \textit{difference} of over 200 hours between the two systesm under central assumptions at 3kt, and a range difference of over 750nm.  These differences are extremely large, representing plausibly double the expected total ranges and endurances of existing modern diesel-electric submarines. Even if such large changes in hotel load and battery capacity are not simultaneously likely these sensitivitiesquite clearly demonstrate how even modest shifts in both hotel load and battery energy density might combine to have very significant impacts on the appearnce of the pump-jet and propeller comparison.  For instance, a hotel load that reduced by 30\%, and a battery energy density increase of 40\% could plausibly have the impact of doubling the nominal differces between the performance of the two propulsive systems.

This sort of extreme sensitivity ought to have a significant bearing on how the overall performance specifications of the submarine are defined, as seemingly incremental changes to hotel load and electrical energy stores are likely to substantially magnified, or muted, by the choice of propulsion system.

@@

\subsection{The importance of cavitation inception}
It is generally considered that the existence of cavitation of any sort will tend to dominate the acoustic signature of any vessel that doesn't produce any machinery noise by relying on all-electric drive as a conventional submarine would when dived (\cite{giles2010}, \cite{harvie1965construction}).  Its quite possible that during the period when a submarine is snorting, or running its diesel engines, these sources of noise would be a more substantial and distinctive acoustic signature than small amounts of cavitation on the propulsion system if it were to occur, owing in particular to the low freqency of diesel engines.

Other sources of noise hydrodynamic noise, such as the turbulence produced in the boundary layer of the flow over the hull, or parts of the propulsor, will still contribute to overall noise and some signature, but it will tend to be significant smaller in magnitude.  This owes to the essentially incompressible nature of water.  Without compression, even very relatively intense votices or tubulent flows will not propagate very much noise, if the water remains perfectly solid.  All of the 'round and round' can't achieve much 'in and out', which is what results in propagated noise.  This obviously doesn't hold true in compressible fluids, such as air, as one would observe noticing the intense noise produced by hand-dryers, aeroplanes, jet engines etc.  It also doesn't hold true \textit{within} a turbulent flow underwater.  If you were to put a sensor inside a turbulent flow, the rapid movements would result in continual and significant pressure fluctuations, hence noise.  But at a distance, these fluctuations tend to substantially cancel out.  The larger the scale of the eddies and vortices, the further noise might propagate, and also the lower the frequencies of resultant noise might be, which makes noise propagate further.  Likewise, larger caviation voids will tend to emit lower frequency sound aswell, as can be observed from the boiling of a kettle.

As a result, when dived, it might be reasonably assumed that the complete elimination of all sources of cavitition will be the primary objective of acoustic signature management.  This is neatly put in \cite[88]{gearhart1966selection}: "The occurence of cavitation in a propulsor results in a degradation of the propulsor operating characteristics if extensive amounts of cavitation exist, and in significant noise when minute amounts occur.  As a result, the performance requirements of a submersible weapon system specify that the occurence or inception of cavitation be avoided below certain operating depths." One might consider that a submarine's 'tactical silent speed' would most probably correspond to the speed at which it can operate safely below the point of any cavitation inception.

@@

Whilst a range of very particular and intricate phenomena may be at work, perhaps the largest and most dominant trend at work will be that pressure differentials across the blade must necessarily increase with increase thrust.  The blade velocity relative to the oncoming stream is also a close indicator, which will be closely related for an open propeller.  Consequently, since the thrust required for self-propulsion increases with velocity squared, the propensity for cavitation occur will necessarily tend to follow this relationship also.  As such, modest changes in speed can have a big impact on the degree of probability of cavitation occurring. Halving your speed will roughly reduce local pressure fluctions by a factor of four. Consequently, even for non-specialised propellers, there is generaly some low speed for which no cavitation occurs, though almost all propellers in major surface vessels experience some cavitation at the speed they are designed to do their work at.

\subsection{The impact of depth on cavitation}
It is also crucial to understand that increasing depth necessarily increases the static pressure throughout a propulsion system.  As such, the speed at which cavitation starts to occur will necessarily be a function of depth.  Approximately every 10m of water adds approximately the equivalent of an atmosphere of pressure.  As such, the static pressure a system experiences at 10m depth will be half that experienced at 30m (since the air in the atmosphere comprises the first atmosphere of pressure).  Consequently, diving deeper will necessarily increase the speeds one can achieve without any cavitation occurring \ref{fig:InceptionDepth.png}.

@@

The interaction of these two relationships can be expressed in a general form in figure \ref{fig:InceptionContours}, which has been generated using an approxiate numeric model. As can be seen here, at very low speeds, modest changes in depths can achieve considerable additional surplus pressure in order to prevent cavitation.  However at much higher speeds, since the pressure diffential scales with the square of the velocity travelled, far greater additional depths are required to achieve sufficient excess static to supress cavitation for a give further speed increase.  Whilst certain system parameters such as the blade area and design (or duct design) will have a significant impact on which curve a particular system lies upon, the overall trend of a parabolic cavitation free "bucket" prevailing at low speeds and large depths will generally be robust.  This is confirmed by the close resemblance between the curve shapes found by an experiment on a torpdo in Figure \ref{fig:InceptionDepth.png}, and the numerical model found in Figure \ref{fig:InceptionContours}.

@@

It is noteworthy also that the vapour pressure of water is an important component in the equations that describe the conditions for cavitation to occur.  It is generally given that when the local static pressure falls below the vapour pressure, cavitation occurs, though known exceptions to this do exist.  Given that temperatures also affects the vapour pressure of water (warmer water is just a little closer to boiling) this does have an impact on the cavitation inception point.  However, at plausible temperatures the impact is small.  The difference between the vapour pressure of water at 10 degrees and 30 degrees is approximately 3 kilopascals.  Diving just 10m deeper adds over 100 kilopascals to static pressure.  In many circumstances, altering your speed by a couple of knots will have an even larger impact on local pressure reductions.  Consequently, the impact of temperature on the inception of cavitation is essentially negligible.  The difference between inception contours in 20 degree and 30 degree water are given by the difference between the black and red-dashed contours in Figure \ref{fig:InceptionContours}.

\subsection{The significance of flow separation}
The addition of a duct will necessarily lead to flow separation occuring on the inside of the duct opening at some very low speeds which correspond to high advance ratios when load remains constant.  Flow separation tends to lead to vortex shedding, which means that inconsistent flows (vortices) are then ingested into the blades of the impeller.  This effect is well known and documented.  A good example is shown on the exterior of an \textbf{accelerating} duct at $J = .8$ in \cite{willemsen2013}.

@@

Whilst at very low power levels or at good depths, it is quite plausible that this doesn't result in any cavitation, which would normally be a likely occurrence if the blades are operating closer to their design power and such such vortices were shed into the intake.  However, given the inherently unpredictable and unsteady nature of such vortex formation and shedding, this phenomenon would increase the risk of cavitation occurring unexpectedly in marginal conditions.  In addition, in the absence of cavitation, the sudden and changing pressure gradients over the blades would dramatically increase the overall pressure fluctuations and turbulence level that occurs within the propulsor.  This would necessarily increase the overall levels of noise generated.  It is plausible that the noise is of sufficiently low level that it is not considered a significant compromise of stealth.  However, it is almost certainly untrue that that pumpjet actually generates less noise overall than a propeller at very low speeds.

@@

It should be noted here that shielding effects might comprise the combination two distinct and seperate physically phenomenon: reflection and damping.  Noise tends to be reflected very effectively off hard, smooth surfaces representing the boundary between mediums of very different density (such as steel and water).  Changes in density which are gradual, often representing soft, squishy, or broken surfaces tend, to have the effect of absorbing sound, where the acoustic energy is converted to heat or degradation of the material, rather than reflecting it.  Given the significance of increased drag on any surfaces within the propulsor, it is highly unlikely that any attempt at damping is made within the pumpjet.  However, it should be noted that some pumpjets have been constructed with a hollow design, in order to reduce weight \parencite{harvie1965construction}, which would reduce the thickness of solid steel, though offer some potential opportunity for damping.  Overall, damping is increased in its effectiveness at low frequencies by allowing it to occur over larger distances, corresponding to the sound's increased wavelength, leading to increased thickness of the transition substance or damping material.  Consequently, effective damping within the propulsor itself is unlikely to be attempted, or limited to medium and higher frequencies, probably suited to reducing noise from normal surface flow turbulence and very early onset cavitation.

@@

It is most likely that little or no damping or absorbing layer is attempted within the body of the shroud, since increasing the duct thickness above hydrodynamically determined optimum profile is likely to have a detrimental impact on efficiency.  Constructing a large components from thing pieces of steel, and still maintaining an optimal shape to the precision required for housing powerful turbomachinery like the impeller is also likely to involve considerable construction risk and complexity.  Hence the shielding effect is most likely to dominated by the reflective characteristics of the duct, which would seek to internally reflect noise back into the turbulent flow, where sound waves would be increasingly convoluted, and somewhat absorbed through the eventual transition to small-scale turbulence and heat.

The degree to which the total noise generated is reflected internally will also be shape-dependent.  A more concave inner surface, will increase the extent of internal reflection.  Such a design would be consistent with higher degrees of diffusion in advance of the impeller, and greater nozzle narrowing after it, which is consistent with higher pressure elevation at the impeller (hence improved cavitation characteristics) but higher negative duct thrust, leading to even poorer efficiencies at low speed.  Substantial alteration of duct shape on account of efficiency considerations at low speed are therefore necessarily going to have some impact on the degree of effective shielding that a duct can afford.

@@

When the submarine is travelling at high speeds, and there is significant flow through the propulsor, the impact of waterspeed through the propulsor might have the additional impact of diminishing the radiation of noise in the forwards direction, in the same way that sound tends tends not to travel well up-wind in air.  However, at very low speeds this effect would be greatly diminished, and possibly not relevant.

It should further be noted that in the underwater environment sound can be made to reflect in a variety of directions simply with differences in temperature, salinity, movement of water bodies, as well as off plenty of nearby underwater objects, including the ocean floor and objects on it.  This phenomenon generally serves to make the art of underwater acoustics complex and unpredictable, and generally contributes to the difficulty of precisely locating a submarine, even when some signature is detected.  However, it also means that direction-dependent reductions signature do not reliably translate to overall reduced probability of detection.

@@

This has an important consequence for shielding for pumpjets at very low speeds because of the occurrence of flow separation and vortex shedding. In the absence of flow separation and cavitation at very low speeds the most substantial sources of sound would be the turbulent flow over surfaces of in the propulsor.  This noise will tend to be relatively white and high in its frequency owing to the very small length-scales of the eddies and vortices in such turbulence.  The steel shroud would be very effective in reflecting such high frequency noise.  Under acceleration or higher loads, the very instances of cavitation woud also be relatively high frequency, as the voids or cavities would be small at first. These sounds might also be very effectively reflected by the shroud, and consequently diminished.

In contrast, flow separation results in the creation of vortices of a dramatically larger scale than occurs with normal turbulence in a boundary later, and hence would tend to generate noise of a much lower frequency.  In addition, as the flow rates slow even further, these vortices are likely to become larger in size, even if lesser in intensity, resulting in further reduction in the frequency of noise generated.  The rate at which the vortices shed from the surface and impact the impeller will also result in another acoustic signature.  Again, at lower flow rates, these vortices will tend to develop and shed more slowly, lowering frequencies further.

@@

The foregoing consideration suggests that the acoustic signature of a jet-powered submarine might not reduce steadily with speed all the way to stationary movement.  It is plausible that there could be some elevated propulsion signature at very low speeds due to flow separation, and that the submarine's optimal acoustic signature when moving is actually at some slightly higher speed, where smoother flow through the propulsor can be achieved.  However, given that the amplitude, frequency, and combined impact of those two characteristics cannot be accurately evaluated without substantial further work and more detailed system knowledge, it is also possible that the increase in other noise (including hull flow noise and other machinery) at elevated speeds offsets the improvement in the smoothness of the pumpjet's operation.  It is still quite likely the overall level of noise radiated by the pumpjet, including under conditions when flow separation occurs, is still considered to be sufficiently low in amplitude as to not be of significant tactical consequence.  However it is highly unlikely that the accoustic signature of the pumpjet is actually superior to that of an open propeller at very low speeds, when cavitation occurs for neither system, but substantially increased turbulence including flow separation must neccessarily occur for the pumpjet.

\section{Conclusion}

The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.  Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.

Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a confrontation between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."  No further explanation of this superiority has been provided in detail, but it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{cameron2016}.

The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.  The conclusion of this report is that the following statement can be made with confidence:
\n
\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.  This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
\n

Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.  One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence\parencite{davies2017}.  This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt. During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.  Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.  Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.
